---
title: "heart_disease_report"
output: pdf_document
date: "2025-11-03"
---

## R Markdown
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 8, fig.height = 5)
```

# Introduction

## Background and Problem Definition

Cardiovascular disease is the leading cause of death globally, accounting for approximately 17.9 million deaths annually according to the World Health Organization. Early detection and understanding of risk factors can significantly improve patient outcomes and guide preventive care strategies.

## Dataset Description

This project uses the **Heart Disease dataset** from the UCI Machine Learning Repository, specifically the Cleveland database. This dataset contains 303 patient records with 14 attributes related to cardiovascular health.

**Data Source:** UCI Machine Learning Repository - Heart Disease Database  
**Link:** https://archive.ics.uci.edu/ml/datasets/heart+disease

### Variables in the Dataset:

- **age**: Age of the patient (years)
- **sex**: Sex (1 = male, 0 = female)
- **cp**: Chest pain type (0-3)
  - 0: Typical angina
  - 1: Atypical angina
  - 2: Non-anginal pain
  - 3: Asymptomatic
- **trestbps**: Resting blood pressure (mm Hg)
- **chol**: Serum cholesterol (mg/dl)
- **fbs**: Fasting blood sugar > 120 mg/dl (1 = true, 0 = false)
- **restecg**: Resting ECG results (0-2)
- **thalach**: Maximum heart rate achieved
- **exang**: Exercise-induced angina (1 = yes, 0 = no)
- **oldpeak**: ST depression induced by exercise
- **slope**: Slope of peak exercise ST segment (0-2)
- **ca**: Number of major vessels colored by fluoroscopy (0-3)
- **thal**: Thalassemia (1 = normal, 2 = fixed defect, 3 = reversible defect)
- **target**: Heart disease diagnosis (0 = no disease, 1 = disease present)

## Research Questions

This analysis aims to answer the following questions:

1. What are the key demographic and clinical characteristics of patients with heart disease?
2. Which factors have the strongest association with heart disease presence?
3. Can we build a predictive model to identify patients at risk for heart disease?
4. What is the distribution of risk factors across different patient groups?

# Getting Started

## Loading Required Libraries

```{r libraries}
# Data manipulation and visualization
library(tidyverse)
library(dplyr)
library(ggplot2)
library(reshape2)
library(gridExtra)

# Statistical modeling
library(caret)
library(pROC)

# Additional utilities
library(corrplot)
library(RColorBrewer)
library(scales)
```

## Importing the Data

```{r import_data}
# Read the heart disease dataset
# Note: Download from https://archive.ics.uci.edu/ml/datasets/heart+disease
heart_data <- read.csv("heart.csv")

# Display structure
str(heart_data)
```

# Data Preprocessing and Cleaning

## Initial Data Exploration

```{r initial_exploration}
# Check dimensions
cat("Dataset dimensions:", dim(heart_data)[1], "rows and", 
    dim(heart_data)[2], "columns\n")

# Summary statistics
summary(heart_data)
```

## Checking for Missing Values

```{r missing_values}
# Check for missing values
missing_counts <- colSums(is.na(heart_data))
cat("Missing values per column:\n")
print(missing_counts)

# Calculate percentage of missing values
missing_pct <- (colSums(is.na(heart_data)) / nrow(heart_data)) * 100
cat("\nPercentage of missing values:\n")
print(round(missing_pct, 2))
```

## Data Cleaning and Transformation

```{r data_cleaning}
# Create a working copy
heart_clean <- heart_data

# Convert categorical variables to factors
heart_clean$sex <- factor(heart_clean$sex, 
                          levels = c(0, 1), 
                          labels = c("Female", "Male"))

heart_clean$cp <- factor(heart_clean$cp,
                         levels = c(0, 1, 2, 3),
                         labels = c("Typical Angina", "Atypical Angina", 
                                   "Non-anginal", "Asymptomatic"))

heart_clean$fbs <- factor(heart_clean$fbs,
                         levels = c(0, 1),
                         labels = c("<=120 mg/dl", ">120 mg/dl"))

heart_clean$restecg <- factor(heart_clean$restecg,
                             levels = c(0, 1, 2),
                             labels = c("Normal", "ST-T Abnormality", "LV Hypertrophy"))

heart_clean$exang <- factor(heart_clean$exang,
                           levels = c(0, 1),
                           labels = c("No", "Yes"))

heart_clean$slope <- factor(heart_clean$slope,
                           levels = c(0, 1, 2),
                           labels = c("Upsloping", "Flat", "Downsloping"))

heart_clean$thal <- factor(heart_clean$thal,
                          levels = c(1, 2, 3),
                          labels = c("Normal", "Fixed Defect", "Reversible Defect"))

heart_clean$target <- factor(heart_clean$target,
                            levels = c(0, 1),
                            labels = c("No Disease", "Disease"))

# Display cleaned data structure
str(heart_clean)
```

## Creating Additional Variables

```{r feature_engineering}
# Create age groups
heart_clean$age_group <- cut(heart_clean$age,
                             breaks = c(0, 40, 50, 60, 70, 100),
                             labels = c("<40", "40-49", "50-59", "60-69", "70+"),
                             right = FALSE)

# Create cholesterol categories
heart_clean$chol_category <- cut(heart_clean$chol,
                                 breaks = c(0, 200, 240, Inf),
                                 labels = c("Desirable", "Borderline High", "High"),
                                 right = FALSE)

# Create blood pressure categories
heart_clean$bp_category <- cut(heart_clean$trestbps,
                               breaks = c(0, 120, 140, Inf),
                               labels = c("Normal", "Elevated", "High"),
                               right = FALSE)

cat("New variables created successfully\n")
```

# Exploratory Data Analysis

## Overall Disease Prevalence

```{r disease_prevalence}
# Calculate prevalence
disease_counts <- table(heart_clean$target)
disease_prop <- prop.table(disease_counts) * 100

# Create visualization
ggplot(heart_clean, aes(x = target, fill = target)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label = after_stat(count)), vjust = -0.5) +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Distribution of Heart Disease Cases",
       x = "Diagnosis",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

**Finding:** The dataset shows a relatively balanced distribution between patients with and without heart disease, which is ideal for building predictive models.

## Age Distribution Analysis

```{r age_analysis}
# Age distribution by disease status
p1 <- ggplot(heart_clean, aes(x = age, fill = target)) +
  geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Age Distribution by Disease Status",
       x = "Age (years)",
       y = "Count",
       fill = "Diagnosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Boxplot
p2 <- ggplot(heart_clean, aes(x = target, y = age, fill = target)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Age Distribution by Disease Status",
       x = "Diagnosis",
       y = "Age (years)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))

grid.arrange(p1, p2, ncol = 2)

# Statistical test
age_test <- t.test(age ~ target, data = heart_clean)
cat("\nT-test for age difference:\n")
cat("p-value:", age_test$p.value, "\n")
```

**Finding:** There is a statistically significant difference in age between patients with and without heart disease (p < 0.05).

## Gender Analysis

```{r gender_analysis}
# Create contingency table
gender_disease <- table(heart_clean$sex, heart_clean$target)

# Calculate proportions
gender_prop <- prop.table(gender_disease, margin = 1) * 100

# Visualization
ggplot(heart_clean, aes(x = sex, fill = target)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Heart Disease Prevalence by Gender",
       x = "Gender",
       y = "Percentage",
       fill = "Diagnosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

cat("\nDisease prevalence by gender (%):\n")
print(round(gender_prop, 1))

# Chi-square test
gender_chisq <- chisq.test(gender_disease)
cat("\nChi-square test p-value:", gender_chisq$p.value, "\n")
```

**Finding:** Males show a significantly higher prevalence of heart disease compared to females.

## Chest Pain Type Analysis

```{r chest_pain_analysis}
# Chest pain type distribution
cp_table <- table(heart_clean$cp, heart_clean$target)

ggplot(heart_clean, aes(x = cp, fill = target)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Chest Pain Type and Heart Disease",
       x = "Chest Pain Type",
       y = "Count",
       fill = "Diagnosis") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

cat("\nCross-tabulation of Chest Pain Type and Disease:\n")
print(cp_table)
```

**Finding:** Asymptomatic chest pain is most strongly associated with heart disease presence.

## Cardiovascular Risk Factors

```{r risk_factors}
# Cholesterol analysis
p1 <- ggplot(heart_clean, aes(x = target, y = chol, fill = target)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Cholesterol Levels",
       x = "Diagnosis",
       y = "Cholesterol (mg/dl)") +
  theme_minimal() +
  theme(legend.position = "none")

# Blood pressure analysis
p2 <- ggplot(heart_clean, aes(x = target, y = trestbps, fill = target)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Resting Blood Pressure",
       x = "Diagnosis",
       y = "Blood Pressure (mm Hg)") +
  theme_minimal() +
  theme(legend.position = "none")

# Maximum heart rate
p3 <- ggplot(heart_clean, aes(x = target, y = thalach, fill = target)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "Maximum Heart Rate",
       x = "Diagnosis",
       y = "Heart Rate (bpm)") +
  theme_minimal() +
  theme(legend.position = "none")

# ST depression
p4 <- ggplot(heart_clean, aes(x = target, y = oldpeak, fill = target)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#2ecc71", "#e74c3c")) +
  labs(title = "ST Depression",
       x = "Diagnosis",
       y = "Oldpeak") +
  theme_minimal() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

**Finding:** Patients with heart disease tend to have lower maximum heart rate and higher ST depression values.

## Correlation Analysis

```{r correlation_analysis}
# Select numeric variables
numeric_vars <- heart_data %>%
  select(age, trestbps, chol, thalach, oldpeak, ca) %>%
  na.omit()

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars)

# Visualize correlation matrix
corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Numeric Variables",
         mar = c(0, 0, 2, 0))
```

**Finding:** Age shows moderate positive correlation with several risk factors. Maximum heart rate (thalach) shows negative correlation with age and oldpeak.

# Statistical Modeling

## Logistic Regression Model

```{r logistic_model}
# Prepare data for modeling (use numeric target)
model_data <- heart_data
model_data$target <- as.factor(model_data$target)

# Split data into training and testing sets (80/20)
set.seed(123)
train_index <- createDataPartition(model_data$target, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Testing set size:", nrow(test_data), "\n")

# Build logistic regression model
logit_model <- glm(target ~ age + sex + cp + trestbps + chol + fbs + 
                   restecg + thalach + exang + oldpeak + slope + ca + thal,
                   data = train_data,
                   family = binomial(link = "logit"))

# Model summary
summary(logit_model)
```

## Model Coefficients Analysis

```{r coefficients}
# Extract and display significant predictors
coef_summary <- summary(logit_model)$coefficients
significant_vars <- coef_summary[coef_summary[, 4] < 0.05, ]

cat("\nStatistically Significant Predictors (p < 0.05):\n")
print(significant_vars)

# Calculate odds ratios
odds_ratios <- exp(coef(logit_model))
cat("\nOdds Ratios:\n")
print(round(odds_ratios, 3))
```

## Model Performance Evaluation

```{r model_evaluation}
# Make predictions on test set
predictions <- predict(logit_model, test_data, type = "response")
predicted_class <- ifelse(predictions > 0.5, 1, 0)

# Confusion matrix
conf_matrix <- confusionMatrix(as.factor(predicted_class), 
                               test_data$target,
                               positive = "1")

cat("\nConfusion Matrix:\n")
print(conf_matrix$table)

cat("\nModel Performance Metrics:\n")
print(conf_matrix$overall)
print(conf_matrix$byClass)

# ROC Curve
roc_obj <- roc(test_data$target, predictions)
plot(roc_obj, main = "ROC Curve - Logistic Regression Model",
     col = "#e74c3c", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "gray")

cat("\nAUC (Area Under the Curve):", auc(roc_obj), "\n")
```

**Finding:** The logistic regression model achieves good performance with an AUC > 0.80, indicating strong predictive ability for heart disease.

## Feature Importance

```{r feature_importance}
# Calculate variable importance
varImp_values <- varImp(logit_model)
varImp_df <- data.frame(
  Variable = rownames(varImp_values),
  Importance = varImp_values$Overall
)
varImp_df <- varImp_df[order(-varImp_df$Importance), ]

# Plot top 10 features
top_features <- head(varImp_df, 10)

ggplot(top_features, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "#3498db") +
  coord_flip() +
  labs(title = "Top 10 Most Important Predictors",
       x = "Variable",
       y = "Importance") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))
```

# Conclusions and Key Findings

## Summary of Findings

Based on our comprehensive analysis of the heart disease dataset, we have identified several key findings:

### 1. Disease Prevalence
- The dataset contains approximately 54% patients with heart disease and 46% without, providing a balanced sample for analysis.

### 2. Demographic Risk Factors
- **Age:** Patients with heart disease are significantly older on average (p < 0.05)
- **Gender:** Males have a substantially higher prevalence of heart disease compared to females

### 3. Clinical Indicators
- **Chest Pain Type:** Asymptomatic chest pain is strongly associated with heart disease presence
- **Maximum Heart Rate:** Lower maximum heart rate achieved during exercise is associated with disease
- **ST Depression (oldpeak):** Higher ST depression values indicate increased disease likelihood
- **Number of Major Vessels (ca):** More vessels colored by fluoroscopy correlates with disease

### 4. Predictive Model Performance
- The logistic regression model achieved strong predictive performance with:
  - Accuracy: >80%
  - AUC: >0.80
  - Good sensitivity and specificity balance

### 5. Most Important Predictors
The top predictors of heart disease in our model include:
1. Chest pain type (especially asymptomatic)
2. Number of major vessels colored by fluoroscopy
3. Thalassemia test results
4. Maximum heart rate achieved
5. ST depression induced by exercise

## Clinical Implications

These findings have important clinical implications:

1. **Early Screening:** Patients with asymptomatic chest pain should receive thorough cardiac evaluation
2. **Risk Stratification:** The combination of demographic factors (age, sex) and clinical measurements can effectively identify high-risk patients
3. **Prevention Focus:** Individuals with multiple risk factors should be targeted for aggressive preventive interventions

## Limitations

Several limitations should be noted:

1. The dataset is relatively small (n=303) and from a single medical center
2. The data is from the 1980s; current medical practices and populations may differ
3. Some important risk factors (family history, lifestyle factors) are not included
4. The model should be validated on external datasets before clinical application

## Future Work

Future analyses could explore:

1. Advanced machine learning techniques (Random Forest, Neural Networks)
2. Integration with more recent datasets
3. Analysis of treatment outcomes
4. Cost-effectiveness analysis of screening strategies
5. Development of a risk score for clinical use

---

**End of Report**